---
title: "THRA Isoform 1 and Isoform 2 Relative Expression"
output:
  html_document:
    toc: True
    toc_float: True
    df_print: paged
---
INTRODUCTION to the Experiment

	A total of 148 solid tissue samples representing 20 organs were taken from post-mortal human healthy donors killed in road accidents no later than 36 hours after death. Prior to RNA extraction solid normal tissue samples were stored either in RNAlater or as Formalin-Fixed Paraffin-Embedded (FFPE) blocks. Blood and bone marrow samples were taken from 6 and 11 healthy volunteers respectively and subjected immediately to RNA extraction. Following ribosomal RNA depletion and RNA libraries construction transcription profiles of the samples were obtained using Illumina HiSeq 3000.
First-Stranded single-end reads. 50bp. Read counts of THRA exon 9a (Chr17:40089333) and exon 9b (Chr17:40089334)

```{r}
## #load libraries
library("DESeq2")
library(tidyverse)
library(ggrepel)
library(dplyr)
library("plotly")



```

```{r}
### Load gene(/transcript) count matrix and labels

countData <- as.matrix(read.csv("gene_count_matrix.csv", row.names="gene_id"))
colData <- read.delim("metadata_GEO_GSE120795.txt", sep=",",row.names="Run")

colData

### countData column names need to be edited. Only the Run number should be used as a name to match colData
colnames(countData) = str_remove(colnames(countData), '_G.*')  ## the . is essential to disregard the next _ characters


### Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colData) %in% colnames(countData))
countData <- countData[, rownames(colData)]
all(rownames(colData) == colnames(countData))

```
```{r}
### Create a DESeqDataSet from count matrix and labels
dds <- DESeqDataSetFromMatrix(countData = countData, 
                                colData = colData, design =  ~ source_name)
```
```{r}
###

colData(dds)
head( counts(dds), 5 )
```
```{r}
### Principal component analysis to detect outliers in the replicates
vsd <- vst(dds, blind=FALSE)
sampleDists <- dist(t(assay(vsd)))

pcaData <- plotPCA(vsd, intgroup=c("source_name"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

#plotPCA(vsd, intgroup=c("source_name"))

x_max_PCA <- max(pcaData$PC1)
x_min_PCA <- min(pcaData$PC1)
y_max_PCA <- max(pcaData$PC2)
y_min_PCA <- min(pcaData$PC2)


### Plot one PCA containing all the samples

ggplot(pcaData, aes(PC1, PC2, color=source_name)) +
  geom_point(size=2.5) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed(xlim = c(x_min_PCA, x_max_PCA), ylim = c(y_min_PCA, y_max_PCA))+
  theme_light(base_size = 12) 
   # geom_encircle(data = pcaData[pcaData$source_name == "Kidney", ], aes(x=PC1, y=PC2), show.legend = TRUE)    # draw circles, needs library(ggalt) and library(ggfortify)
  
ggsave("PCA_total.png", width = 32, height = 20, units = "cm")
ggplotly()



### Plot all the samples in one big grid
PCA_single <- ggplot(pcaData, aes(PC1, PC2, color=source_name)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed(xlim = c(x_min_PCA, x_max_PCA), ylim = c(y_min_PCA, y_max_PCA)) + 
  theme(legend.position="none") +
  theme_light(base_size = 12) +
  geom_text_repel(aes(label =  name),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 170,
                  size=2)

PCA_single + facet_wrap(~source_name) + theme(legend.position="none") 

ggsave("PCA_single_samples.png", width = 32, height = 20, units = "cm")


### Plot each sample in one single PCA plot

for (sample in pcaData$source_name ) {

  graph_title <- paste0("PCA_", sample, ".png")
  
  PCA_single %+% subset(pcaData, source_name %in% paste(sample))
  
  
  ggsave(graph_title, width = 32, height = 20, units = "cm", path= "~/OneDrive - Charité - Universitätsmedizin Berlin/Schuelke_Lab/EG15_RNA_Seq/GEO_GSE120795/PCA_single_outputs")
  
  
  }


```


















```{r}
#set the reference level
dds$Condition <- relevel(dds$Condition, ref = "higher_THRA1")

### run DESeq2 and calculate the results
dds <- DESeq(dds)
res <- results(dds)
res

```
```{r}
### Log fold change shrinkage for visualization and ranking
resultsNames(dds)

resLFC <- lfcShrink(dds, coef="Condition_lower_THRA1_vs_higher_THRA1", type="apeglm")
resLFC
```
```{r}
### p-values and adjusted p-values
resOrdered <- res[order(res$padj), ]
summary(res)

```

```{r}
### How many adjusted p-values were less than 0.05?
sum(res$padj < 0.05, na.rm=TRUE)
```

```{r}
plotDispEsts( dds )

plotMA(resLFC, ylim=c(-10,10), alpha= 0.05, main= "MA plot with alpha=0.05")

```

```{r}
### plotting the genecounts
plotCounts(dds, gene="ENSG00000126351.13|THRA", intgroup="Condition")

```
```{r}
### exporting the results
resSig <- subset(resOrdered, padj < 0.05)
resSig

write.csv(as.data.frame(resSig), 
          file="Low_THRA1_vs_high_as_ref_results.csv")
```

```{r}
library("pheatmap")
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("Condition", "Tissue")])
ntd <- normTransform(dds)
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
```

 
 Another use of the transformed data is sample clustering. Here, we apply the dist function to the transpose of the transformed count matrix to get sample-to-sample distances.
 A heatmap of this distance matrix gives us an overview over similarities and dissimilarities between samples. We have to provide a hierarchical clustering hc to the heatmap function based on the sample distances, or else the heatmap function would calculate a clustering based on the distances between the rows/columns of the distance matrix.
 
```{r}
### clustering

library("RColorBrewer")

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Condition, vsd$Tissue, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
heatmap <-pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

heatmap
ggsave( "Heatmap.pdf", width = 40, height = 25, units = "cm")

```


```{r}
# make screeplot
    ## calculate the variance for each gene
     rv <- rowVars(assay(vsd))

     ## select the ntop genes by variance
     select <- order(rv, decreasing=TRUE)[seq_len(min(500, length(rv)))]
     
     ## perform a PCA on the data in assay(x) for the selected genes
     pca <- prcomp(t(assay(vsd)[select,]))
     
     ## the contribution to the total variance for each component
     percentVar <- pca$sdev^2 / sum( pca$sdev^2 )*100
     
     ##plot the "percentVar"
     scree_plot=data.frame(percentVar)
     scree_plot[,2]<- c(1:6)

     colnames(scree_plot)<-c("variance","component_number")
     ggplot(scree_plot, mapping=aes(x=component_number, y=variance))+geom_bar(stat="identity")+
         ggtitle("Scree plot") +
      theme(legend.position = "right", plot.title = element_text(hjust=0.5)) + 
       ylab("% Variance")
```
 
 
```{r}
### P-values distribution
## P-value histogram plot
res.df <- as.data.frame(res)

## Sort results by adjusted p-values
ord <- order(res.df$padj, decreasing = FALSE)
res.df <- res.df[ord, ]
res.df <- cbind(data.frame(Feature = rownames(res.df)), res.df)
rownames(res.df) <- NULL

## Adjusted p-values histogram plot
ggplot(res.df[!is.na(res.df$padj), ], aes(x = padj)) +
    geom_histogram(alpha=.5, position='identity', bins = 50) +
    labs(title=paste('Histogram of', elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)])) +
    xlab('Adjusted p-values') +
    xlim(c(0, 1.0005))
```

BiocManager::install("ReportingTools")
citation("ReportingTools")
https://bioconductor.org/packages/devel/bioc/vignettes/ReportingTools/inst/doc/rnaseqAnalysis.pdf

```{r}
library(ReportingTools)

des2Report <- HTMLReport(shortName = 'RNAseq_analysis_with_DESeq2_P0.1',
title = 'RNA-seq analysis of differential expression: Tissues expressing high vs low THRA1',
reportDirectory = "./reports")

publish(dds,des2Report, pvalueCutoff=0.1,
annotation.db="org.Mm.eg.db", factor = colData(dds)$Condition,
reportDir="./reports")
finish(des2Report)


```


```{r}
### get reference list for GO
GO_reference_list <- dplyr::select(res.df, Count=baseMean,  GeneID= Feature, FDR="padj", Log2FoldChange="log2FoldChange" ) %>%
   drop_na() %>%
 #  separate_rows(GeneID, sep="\\|") %>% #splits multiple IDs on same row and keeps the left option
  #separate_rows(GeneID, sep="\\.") %>% #splits multiple IDs on same row and keeps the left option
  separate(GeneID, "ENSEMBL_ID", sep="\\.", remove=FALSE, extra="warn", fill="right") %>% #splits multiple IDs on same row and keeps the right "ENSEMBLID"
  group_by(GeneID) %>%
   arrange(FDR, .by_group=T) %>% #groups by Gene name and sorts them by FDR
   top_n(-1, FDR) %>% # selects for each gene the one with lowest FDR
   distinct()


write.csv(GO_reference_list$ENSEMBL_ID, 
          file="GO_reference_list.csv")

```


```{r}
# Volcano Plot of results
library(EnhancedVolcano)
#library("biomaRt")
#ensembl <- useMart("ensembl",dataset="hsapiens_gene_ensembl") #retrieves human database GRCh38.p13
#ensembl_gene_names <- getBM(mart=ensembl, attributes = c("ensembl_gene_id", "external_gene_name"), filters = "chromosome_name", values=1:22)

#ensembl_gene_names <- ensembl_gene_names %>%
 #  rename(ENSEMBL_ID=ensembl_gene_id, Gene=external_gene_name)

Volcano_list <- dplyr::select(res.df, GeneID= Feature, FDR="padj", Log2FoldChange="log2FoldChange" ) %>%
   drop_na() %>%
 #  separate_rows(GeneID, sep="\\|") %>% #splits multiple IDs on same row and keeps the left option
  #separate_rows(GeneID, sep="\\.") %>% #splits multiple IDs on same row and keeps the left option
  separate(GeneID, "ENSEMBL_ID", sep="\\.", remove=FALSE, extra="warn", fill="right") %>% #splits multiple IDs on same row and keeps the right "ENSEMBLID"
  group_by(GeneID) %>%
   arrange(FDR, .by_group=T) %>% #groups by Gene name and sorts them by FDR
   top_n(-1, FDR) %>% # selects for each gene the one with lowest FDR
   distinct()


Volcano_list_genes <- Volcano_list %>%
  separate(GeneID, c("delete","Gene"), sep="\\|", remove=FALSE, extra="drop", fill="right")%>% #splits multiple IDs on same row and keeps the right
  dplyr::select(-delete)
  
Volcano_list_labels <- dplyr::filter(Volcano_list_genes, FDR<0.05)
write.csv(as.data.frame(Volcano_list_labels), 
          file="Significant_genes_results_0.05.csv")


FC <- 0.5
p <- 0.05

keyvals <- rep('grey75', nrow(Volcano_list_genes))
names(keyvals) <- rep('Not Significant', nrow(Volcano_list_genes))

#keyvals[which(abs(Volcano_list$Log2FoldChange) > FC & Volcano_list$FDR > p)] <- 'grey50'
#names(keyvals)[which(abs(Volcano_list$Log2FoldChange) > FC & Volcano_list$FDR > p)] <- 'Log2FoldChange'

#keyvals[which(abs(Volcano_list$Log2FoldChange) < FC & Volcano_list$FDR < p)] <- 'grey25'
#names(keyvals)[which(abs(Volcano_list$Log2FoldChange)  < FC & Volcano_list$FDR < p)] <- '-Log10Q'

keyvals[which(Volcano_list_genes$Log2FoldChange < -FC & Volcano_list_genes$FDR < p)] <- 'blue2'
names(keyvals)[which(Volcano_list_genes$Log2FoldChange  < -FC & Volcano_list_genes$FDR < p)] <- 'Down-regulated'

keyvals[which(Volcano_list_genes$Log2FoldChange > FC & Volcano_list_genes$FDR < p)] <- 'red2'
names(keyvals)[which(Volcano_list_genes$Log2FoldChange > FC & Volcano_list_genes$FDR < p)] <- 'Up-regulated'

unique(keyvals)
unique(names(keyvals))


volcano_plot <- EnhancedVolcano(Volcano_list_genes,
    lab = Volcano_list_genes$Gene,
    x = 'Log2FoldChange',
    y = 'FDR',
    #selectLab = GBA_final_list$gene ,
    #selectLab = c('ENSG00000153113','ENSG00000142192', 'ENSG00000168002', 'ENSG00000087460', 'ENSG00000148935', 'ENSG00000138162'), #for specific genes 
    pCutoff = 0.05,
    FCcutoff = 0.5,
    xlim = c(-5.5, 5.5),
    ylim = c(0, -log10(10e-12)),
    colCustom = keyvals,
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10] ~ italic(Padj)),
    pointSize = 2.5,
    #labSize = 3.0,
    #shape = c(6, 6, 19, 16),
    title = "Expression pattern of tissues expressing less THRA1",
    subtitle = "Differential Expression Analysis, High THRA1 expressing tissues as reference",
    caption = "Fold Change cutoff = 0.5; Padj (FDR) cutoff = 0.05",
    #legend=c('NS','Log (base 2) fold-change','P value', 'P value & Log (base 2) fold-change'),
    #legend=c('','','P value', 'P value & Log (base 2) fold-change'),
    legendPosition = "right",
    legendLabSize = 15,
    legendIconSize = 10,
    col = c("grey30", "grey30", "royalblue", "red2"),
    colAlpha = 0.8,
    drawConnectors = TRUE,
    widthConnectors = 0.6,
    colConnectors = 'grey50',
    boxedLabels=T,labSize =3)
   # hline = c(10e-8), #adds cutoff lines

volcano_plot


ggsave("Volcano_plot.pdf", width = 40, height = 25, units = "cm")
```





