---
title: "THRA Isoform 1 and Isoform 2 Relative Expression"
output:
  html_document:
    toc: True
    toc_float: True
    df_print: paged
---

INTRODUCTION to the Experiment

	A total of 148 solid tissue samples representing 20 organs were taken from post-mortal human healthy donors killed in road accidents no later than 36 hours after death. Prior to RNA extraction solid normal tissue samples were stored either in RNAlater or as Formalin-Fixed Paraffin-Embedded (FFPE) blocks. Blood and bone marrow samples were taken from 6 and 11 healthy volunteers respectively and subjected immediately to RNA extraction. Following ribosomal RNA depletion and RNA libraries construction transcription profiles of the samples were obtained using Illumina HiSeq 3000.
First-Stranded single-end reads. 50bp. Read counts of THRA exon 9a (Chr17:40089333) and exon 9b (Chr17:40089334)

```{r}
#load libraries
library(tidyverse)
library(readr)
library(ggplot2)
library(plotly)
library(matrixStats)
library(ggrepel)
library(scales)
library(readxl)
library(dplyr)

```


```{r}
## set paths for output figure
path_plots <- "~/OneDrive - Charité - Universitätsmedizin Berlin/Schuelke_Lab/EG15_RNA_Seq/GEO_GSE120795/THRA_isoform_Expression_Coverage"

## load metadata file

metadata <- read.delim("metadata_GEO_GSE120795.txt", sep=",")
metadata_tissues <- select(metadata, Run, "Sample" =source_name)


##Load bedtools outputs
file_links_bedtools <- list.files(path= "~/OneDrive - Charité - Universitätsmedizin Berlin/Schuelke_Lab/EG15_RNA_Seq/GEO_GSE120795/bedtools_outputs" , pattern = "*.txt", full.names=T)


# initialize an empty dataframe
data_bedtools <- data_frame("V1"=character(),
                   "V2"=integer(),
                   "V3"=integer(),
                   "V4"=character(),
                   "V5"=double(),
                   "Run"=character()) 

for (x in file_links_bedtools) {
  table_sample <- read.delim(x, header = FALSE) ## read table
  basename_sample <- basename(x)
  table_sample <- mutate(table_sample, "Run"=basename_sample)
  sum_column <- table_sample$V5
  
  # if statement to filter out samples that have 0 coverage in both exon 9a and 9b
  if (sum(sum_column)  > 4) {
    data_bedtools <- bind_rows(data_bedtools, table_sample)
  }
    
  
}

#extract the SSR number and merge with the Sample name
data_bedtools <- data_bedtools %>%
  separate("Run","Run", "_") %>%
  left_join(metadata_tissues, by="Run")

data_bedtools

##Load mosdepth outputs

file_links_mosdepth <- list.files(path= "~/OneDrive - Charité - Universitätsmedizin Berlin/Schuelke_Lab/EG15_RNA_Seq/GEO_GSE120795/mosdepth_outputs" , pattern = "*.bed", full.names=T)


# initialize an empty dataframe
data_mosdepth <- data_frame("V1"=character(),
                   "V2"=integer(),
                   "V3"=integer(),
                   "V4"=character(),
                   "V5"=double(),
                   "Run"=character()) 

for (x in file_links_mosdepth) {
  table_sample <- read.delim(x, header = FALSE) ## read table
  basename_sample <- basename(x)
  table_sample <- mutate(table_sample, "Run"=basename_sample)
  treshold_value <- filter(table_sample, V4 != "THRA") %>%
    select(V5)
 # treshold_value <- treshold_value$V5
  
  # if statement to filter out samples that have 0 coverage in both exon 9a and 9b
  if (sum(treshold_value)  > 4) {  ## generally should be at least 4
    data_mosdepth <- bind_rows(data_mosdepth, table_sample)
  }
    
  
}

#extract the SSR number and merge with the Sample name
data_mosdepth <- data_mosdepth %>%
  separate("Run","Run", "_")%>%
  left_join(metadata_tissues, by="Run")

data_mosdepth

```

```{r}
## Let's rearrange the data in a useful way
## I will keep track of Run as the replicate identifier
data_mosdepth <- data_mosdepth %>%
  select(Sample, Run, Isoform = V4, Reads_count =V5) %>%
  spread(key="Isoform", value="Reads_count") %>%
  mutate("Package" = as.factor("mosdepth")) 

data_mosdepth

data_bedtools <- data_bedtools %>%
  select(Sample,Run, Isoform = V4, Reads_count =V5) %>%
  spread(key="Isoform", value="Reads_count")%>%
  mutate("Package" = as.factor("bedtools")) %>%
  filter(Run %in% data_mosdepth$Run) ## makes sure we take only samples that were filtered already

data_bedtools

```

```{r}
## Let's merge the two dataframes into one

dataset <- full_join(data_mosdepth, data_bedtools) 

dataset
```



```{r}
## Let's calkculate THRA1 (counts of 9b) and THRA2 (9a-9b)

dataset <- dataset %>%
  mutate("THRA1"= dataset$`9b`) %>%
  mutate("THRA2"=dataset$`9a`-dataset$`9b`) %>%
  rename("Read_counts_9a" = "9a") %>%
  rename("Read_counts_9b" = "9b") %>%
  relocate(Package, .after=Sample)

dataset

```
```{r}
## Let's add the final calculations

dataset_final <- dataset %>%
  mutate("delta_A1vsA2" = THRA1 - THRA2) %>%
  mutate("THRA1_Percentage" = round(THRA1/Read_counts_9a*100)) %>%
  mutate("THRA2_Percentage" = round(100-THRA1_Percentage)) %>%
  mutate("THRA1_higher" = THRA1 > THRA2) %>%
  mutate("delta_percentage" = THRA1_Percentage - THRA2_Percentage)

dataset_final
  
```






```{r}
### Lets compare the read counts of exon 9a between the 2 tools
graph_data_exon9a <- dataset_final %>%
  select(Sample, Package, Read_counts_9a)


ggplot(graph_data_exon9a, aes(x = Sample, y = Read_counts_9a, fill = Package)) +
  geom_boxplot( width=0.5, position=position_dodge(width=0.7)) +
  scale_y_continuous("Read counts") +
  scale_fill_manual("", values = c("bedtools" = "darksalmon", "mosdepth" = "#56B4E9")) +
  ggtitle("Tools comparison - Exon 9a ") +
  theme_light(base_size = 12) +
  scale_x_discrete(guide = guide_axis(angle = -35)) +
  theme(axis.title.x=element_blank())

#ggplotly()

## save plot
ggsave("Read_counts_9a_tool_comparison.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```




```{r}
### Lets compare the read counts of exon 9b between the 2 tools
graph_data_exon9b <- dataset_final %>%
  select(Sample, Package, Read_counts_9b)


ggplot(graph_data_exon9b, aes(x = Sample, y = Read_counts_9b, fill = Package)) +
  geom_boxplot( width=0.5, position=position_dodge(width=0.7)) +
  scale_y_continuous("Read counts") +
  scale_fill_manual("", values = c("bedtools" = "darksalmon", "mosdepth" = "#56B4E9")) +
  ggtitle("Tools comparison - Exon 9b ") +
  theme_light(base_size = 12) +
  scale_x_discrete(guide = guide_axis(angle = -35)) +
  theme(axis.title.x=element_blank())

#ggplotly()
## save plot
ggsave("Read_counts_9b_tool_comparison.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```




```{r}
### heatmap with continuos delta THRA1vsA2 using mosdepth
heatmap_data_mosdepth <- filter(dataset_final, Package == "mosdepth") %>%
  group_by(Sample, Package) %>%
  summarize(delta_A1vsA2=round(mean(delta_A1vsA2)))

y_max_mosdepth <- max(heatmap_data_mosdepth$delta_A1vsA2)
y_min_mosdepth <- min(heatmap_data_mosdepth$delta_A1vsA2)

ggplot(heatmap_data_mosdepth, aes(1, reorder(Sample,delta_A1vsA2), fill=delta_A1vsA2)) +
  geom_tile()+
  ggtitle("THRA isoform expression pattern (THRA1/THRA2) - mosdepth") +
  theme_light(base_size = 12)+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_gradient2('delta_reads', limits=c(y_min_mosdepth, y_max_mosdepth), breaks = c( y_min_mosdepth, 0, y_max_mosdepth),  low = "#56B4E9", high = "darksalmon", guide="colorbar")

ggsave("Heatmap_THRA1vsA2_mosdepth.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```
```{r}
### heatmap with AUT AUT delta THRA1vsA2 using mosdepth
heatmap_data_mosdepth <- heatmap_data_mosdepth %>%
  mutate("THRA1_higher" = delta_A1vsA2 > 0)

ggplot(heatmap_data_mosdepth, aes(1, reorder(Sample,delta_A1vsA2), fill=THRA1_higher)) +
  geom_tile()+
  ggtitle("THRA isoform expression pattern (THRA1/THRA2) - mosdepth") +
  theme_light(base_size = 12)+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_manual(values = c("TRUE"= "darksalmon", "FALSE" = "#56B4E9"))

ggsave("Heatmap_THRA1vsA2_AUT_AUT_mosdepth.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```

```{r}
### heatmap of percentages using mosdepth
heatmap_data_mosdepth <- filter(dataset_final, Package == "mosdepth") %>%
  group_by(Sample, Package) %>%
  summarize(THRA1_Percentage=round(mean(THRA1_Percentage)))

ggplot(heatmap_data_mosdepth, aes(1, reorder(Sample,THRA1_Percentage), fill=THRA1_Percentage)) +
  geom_tile()+
  ggtitle("THRA isoform expression pattern (THRA1/THRA2) - mosdepth") +
  theme_light(base_size = 12)+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_gradient2('% of THRA1', limits=c(0, 100), breaks = c( 0, 50, 100),  low = "#56B4E9", high = "darksalmon", guide="colorbar",  midpoint = 50)

  ggsave("Heatmap_THRA1vsA2_mosdepth_percentages.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```

```{r}
### Lets plot the percentage in a different way, so that the actual percentage is visible

heatmap_data_percentage <- filter(dataset_final, Package == "mosdepth")%>%
  group_by(Sample) %>%
  summarize(THRA1_Percentage=round(mean(THRA1_Percentage))) %>%
  mutate("THRA1_higher" = THRA1_Percentage >50 ) %>%
  mutate("delta_percentage" = THRA1_Percentage - (100 - THRA1_Percentage))

heatmap_data_percentage <- heatmap_data_percentage[order(heatmap_data_percentage$THRA1_Percentage), ]  # sort
heatmap_data_percentage$Sample <- factor(heatmap_data_percentage$Sample, levels = heatmap_data_percentage$Sample)  # convert to factor to retain sorted order in plot.


# Diverging Barcharts
ggplot(heatmap_data_percentage, aes(x=Sample, y=THRA1_Percentage, label=THRA1_Percentage)) + 
  geom_bar(stat='identity', aes(fill=THRA1_higher), width=.5)  +
   theme_light(base_size = 12)+
  scale_fill_manual(name="Predominant Isoform", 
                    labels = c("THRA1", "THRA2"), 
                    values = c("TRUE"="darksalmon", "FALSE"="#56B4E9")) + 
  scale_y_continuous("% THRA1" ,limits=c(0, 100))+ 
  geom_hline(yintercept=50, linetype="dashed", color = "black")+
  labs(subtitle="Percentage of THRA1 expression over the total of THRA", 
       title= "THRA isoform expression pattern") + 
    theme(axis.title.y=element_blank()) +
  coord_flip()

ggsave("THRA1vsA2_mosdepth_percentages.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )



# Diverging Barcharts
ggplot(heatmap_data_percentage, aes(x=Sample, y=delta_percentage, label=delta_percentage)) + 
  geom_bar(stat='identity', aes(fill=THRA1_higher), width=.5)  +
   theme_light(base_size = 12)+
  scale_fill_manual(name="Predominant Isoform", 
                    labels = c("THRA1", "THRA2"), 
                    values = c("TRUE"="darksalmon", "FALSE"="#56B4E9")) + 
  scale_y_continuous("Delta THRA1-THRA2 in %" ,limits=c(-100, 100))+
  labs(subtitle="Delta THRA1-THRA2 expressed in percentage", 
       title= "THRA isoform expression pattern") + 
    theme(axis.title.y=element_blank()) +
  coord_flip()



```

```{r}
### Lets add a graph where the replicates are visible
heatmap_data_percentage_boxplot <- filter(dataset_final, Package == "mosdepth") %>%
  select(Sample, THRA1_Percentage)  %>%
  group_by(Sample) %>%
  mutate(THRA1_Percentage_mean=round(mean(THRA1_Percentage))) %>%
  mutate("THRA1_higher" = THRA1_Percentage_mean >50 ) %>%
  mutate(replicates = sum(Sample == Sample)) %>%  ## counts the repicates (the n number)
  mutate(labels = paste0(Sample, ", n=", replicates)) %>%
  ungroup()

heatmap_data_percentage_boxplot <- heatmap_data_percentage_boxplot[order(heatmap_data_percentage_boxplot$THRA1_Percentage_mean), ]  # sort
heatmap_data_percentage_boxplot <- mutate(heatmap_data_percentage_boxplot, ID = row_number())

# Diverging Barcharts
ggplot(heatmap_data_percentage_boxplot, aes(x=reorder(labels, ID), y=THRA1_Percentage, label=THRA1_Percentage)) + 
  geom_boxplot(aes(fill=THRA1_higher), width=.5)  +
  #  geom_jitter(color="black", size=0.9, alpha=0.9) +  ## optional in case of a lot of replicates to avoid overlapping of dots
   theme_light(base_size = 12)+
  scale_fill_manual(name="Predominant Isoform", 
                    labels = c("THRA1", "THRA2"), 
                    values = c("TRUE"="darksalmon", "FALSE"="#56B4E9")) + 
  scale_y_continuous("% THRA1" ,limits=c(0, 100))+ 
  geom_hline(yintercept=50, linetype="dashed", color = "black")+
  labs(subtitle="Percentage of THRA1 expression over the total of THRA", 
       title= "THRA isoform expression pattern") + 
    theme(axis.title.y=element_blank()) +
  coord_flip()


ggsave("THRA1vsA2_mosdepth_percentages_boxplot.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )

```


```{r}
### heatmap with continuos delta THRA1vsA2 using bedtools
heatmap_data_bedtools <- filter(dataset_final, Package == "bedtools")%>%
  group_by(Sample, Package) %>%
  summarize(delta_A1vsA2=round(mean(delta_A1vsA2)))

y_max_bedtools <- max(heatmap_data_bedtools$delta_A1vsA2)
y_min_bedtools <- min(heatmap_data_bedtools$delta_A1vsA2)

ggplot(heatmap_data_bedtools, aes(1, reorder(Sample,delta_A1vsA2), fill=delta_A1vsA2)) +
  geom_tile()+
  ggtitle("THRA isoform expression pattern (THRA1/THRA2) - bedtools") +
  theme_light(base_size = 12)+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_gradient2('delta_reads', limits=c(y_min_bedtools, y_max_bedtools), breaks = c( y_min_bedtools, 0, y_max_bedtools),  low = "#56B4E9", high = "darksalmon", guide="colorbar")

ggsave("Heatmap_THRA1vsA2_bedtools.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```





```{r}
### heatmap with AUT AUT delta THRA1vsA2 using bedtools
heatmap_data_bedtools <- heatmap_data_bedtools %>%
  mutate("THRA1_higher" = delta_A1vsA2 > 0)

ggplot(heatmap_data_bedtools, aes(1, reorder(Sample,delta_A1vsA2), fill=THRA1_higher)) +
  geom_tile()+
  ggtitle("THRA isoform expression pattern (THRA1/THRA2) - bedtools") +
  theme_light(base_size = 12)+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_manual(values = c("TRUE"= "darksalmon", "FALSE" = "#56B4E9"))

ggsave("Heatmap_THRA1vsA2_AUT_AUT_bedtools.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```



```{r}
###Lets compare the methods
heatmap_data_comparison <- select(dataset_final, Sample, Package, delta_A1vsA2)%>%
  group_by(Sample, Package) %>%
  summarize(delta_A1vsA2=round(mean(delta_A1vsA2)))

ggplot(heatmap_data_comparison, aes(Package, reorder(Sample,`delta_A1vsA2`) , fill=`delta_A1vsA2`)) +
  geom_tile()+
  ggtitle("Delta comparison between methods (THRA1/THRA2)") +
  theme_light(base_size = 12)+
  theme(axis.title.y=element_blank()) +
  theme(axis.title.x=element_blank()) +
  scale_fill_gradient2('delta', breaks = c(min(y_min_bedtools, y_min_mosdepth), 0,  max(y_max_bedtools,y_max_mosdepth )),  low = "#56B4E9", high = "darksalmon", guide="colorbar")


ggsave("Heatmap_methods_comparison.png" , device=png, dpi = 600, path = path_plots, bg = "transparent", width = 20, height = 15, units = "cm" )
```

